Testing the correlates of huge vertebrae number disparity in Australian blindsnakes

### 2024-10-10 manuscript.Rmd
including feedback from Scott

Evolution of the axial skeleton is fundamental to vertebrate body plan evolution, facilitating body form changes and influencing modes of locomotion in mammals [e.g., @taewcharoenSmall2024;@galisFast2014], reptiles [e.g., @vandammeSpeed2002;@molnarMorphological2023], amphibians [e.g., @bonettEvidence2017;@adlerEvolutionary2022] and fish [e.g., @yamadaAdaptive2009;@forkerBendy2021]. Axial skeleton evolution is often studied by counting the number of vertebrae because the developmental mechanisms of segmentation are well characterised. Essentially vertebrae begin as somites, epithelial blocks of mesoderm, which form rhythmically during embryogenesis where the speed of this process relates to the size and number of somites, to produce what will become the axial skeleton [@gomezControl2008;@dequeantSegmental2008;@gomezDevelopmental2009]. Elongate animals are expected to show a positive correlation between maximum adult body length and number of vertebrae, known as pleomerism [@lindseyPleomerism1975]. While pleomerism was considered the norm [@lindseyPleomerism1975;@wakeComparative1966;@pollyTesting2001], more recent studies have identified exceptions to pleomerism rule in moray eels [@reeceEvolutionary2013], large species of snakes [@headDissociation2007;@lindellEvolution1994], and salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].

The lack of pleomerism in some species suggests it is possible to disassociate segment production with body growth, and thus reveals the developmental processes underlying evolutionary patterns. Another clue to disassociation in reptiles can be found in the relationship between the number of ventral scales and precloacal vertebrae – a correlation that is 1:1 in many snakes [@alexanderPattern1966] and so ventral scales have often been used as a proxy for vertebrae number [@lindellEvolution1994;@shineVertebral2000]. However, this 1:1 correlation is absent in file snakes,*Achrochodus*, and some blindsnakes, Typhlopidae, [@alexanderPattern1966;@headDissociation2007], as well as some sea snakes of the genus *Hydrophis* [@vorisDermal1975], suggesting these taxa do not conform to pleomerism.
Snakes are therefore a good model to study axial evolution because they have great diversity in vertebrae number and a wide range of adult body sizes and shapes but without a limb skeleton to confound signals of locomotor adaptation. Recent studies have indicated the number of vertebrae varies predictably with ecological diversification in elapid snakes [@sherrattVertebral2019;@hamptonInterspecific2019;@sherrattMacroevolution2022] and colubrid snakes [@nash-hahnModification2024]. 
The number of vertebrae has also been shown to influence snake locomotion, whereby the number of vertebrae negatively correlates with speed travelled in garter snakes [@kelleyEffects1997]. Despite these recent advances, there have been no study that have quantified the number of vertebrae across all the species in a diverse assemblage to test the prevalence of pleomerism in fossorial snakes.

<!-- [Specific background]-->
Scolecophidians, commonly known in English as "blindsnakes" and "threadsnakes", are characterised by a reduced visual system and highly specialised phenotypes adapted for a strictly fossorial existence [@mirallesMolecular2018;@strongDeconstructing2021;@gowerEyetranscriptome2021]. Scolecophidians comprise five paraphyletic families, including the "core scolecophidians" — a monophyletic clade consisting of Typhlopidae (425 spp), Leptotyphlopidae (143 spp), Gerrhopilidae (29 spp), and Xenotyphlopidae (1 spp) — and the paraphyletic Anomalepididae (20 spp), which is sister to all other "typical snakes" (Alethinophidia) [@mirallesMolecular2018; @burbrinkInterrogating2020; @uetzQuarter2021]. Although all blindsnakes are fossorial, body size and shape vary significantly among species [@tiatragulMorphological2024; @shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993]. Leptotyphlopids are generally considered more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@hedgesLower2008;@webbDietary2001;@webbPreysize1993;@iturriagaNew2021;@tiatragulMorphological2024]. These differences in body size and proportions determine how much push force the snakes can generate when burrowing, which led to the idea that leptyphlopids generally use existing burrows while typhlopids create their own burrow [@herrelBurrowing2021].

@@

We collected body size measurements and vertebral counts from museum specimens examined. Body measurements including snout-to-vent length (SVL), tail lengths (TL) and midbody diameter (MBD) were obtained from @tiatragulMorphological2024. SVL and TL were measured to the nearest 1 mm on a ruler. MBD was measured using a digital vernier calliper (ABS Coolant Proof Caliper, Mitutoyo, Japan) to the nearest 0.1 mm. Total body length (TBL) is the sum of SVL and TL. To quantify body shape for each specimen, we calculated the aspect ratio [also referred to as elongation ratio; *sensu* @wardConvergent2004], which is the quotient of TBL and MBD.

@@

We used the Fiji distribution of ImageJ2 v.1.54j [@schindelinFiji2012;@ruedenImageJ22017] to digitally visualise X-ray images and counted the number of vertebrae using the Cell Counter module. Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023]. We calculated vertebrae ratio by dividing the total number of vertebrae by maximum TBL.

@@

All statistical analyses were performed in R Statistical Environment v 4.4.0 [@rcoreteamLanguage2024]. To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Brownian Motion implemented in the R package *`geomorph`* v.4.0.6 [@adamsGeomorph2023]. First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes. We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum TBL. Next, we tested whether vertebrae ratio is related to the mean vertebrae number and mean aspect ratio. Finally, we tested if variation in vertebrae ratio and vertebrae number is correlated with maximum soil bulk density and mean annual temperature. In all PGLS analyses, statistical significance was assessed by permutation, using 1000 iterations. To visualise the relationship between vertebrae number and intracolumnar vertebral length, for each of the representative species, we plotted the mean vertebrae size (from three repeated measurements) against the vertebra number. Each vertebrae profile was overlaid with a fitted 4° polynomial curve [e.g., @sherrattMacroevolution2022].

@@

In a related study, @tiatragulMorphological2024 found that Australian blindsnake species with smaller absolute body size generally occupy areas that are warmer regions of Australia than while larger species tend to occupy cooler regions. Here, we observe that vertebrae ratio (raw vertebrae number/SVL) is positively correlated with temperature, indicating species that have more vertebrae than expected from their body size are found in warmer places. In his work on rattlesnakes, @klauberRattlesnakes1982  noted that smaller-sized subspecies often retain the same number of vertebrae as their ancestors, which result in the smaller species having relatively greater vertebrae ratio. An explanation for this pattern is that size is often under greater selection than vertebral number, and there is negligible disadvantage for a snake in having too many vertebrae for their body size [@lindellEvolution1994]. This makes sense when we also consider the negative relationship between vertebrae ratio and maximum soil bulk density. A greater number of (shorter) vertebrae for body size may afford greater flexibility [@jayneKinematics1986;@bergmannVertebral2012;@johnsonAdaptive1955] that enables these species to navigate in pre-existing burrows or termite galleries, typical across northern Australia, without exerting force to dig their own burrow. @tiatragulMorphological2024 also found species that occupy areas with more difficult-to-penetrate soil are usually narrow-and-long rather than robust-and-wide species that occupy areas with easier-to-penetrate soil. Narrow-and-wide phenotypes are common among leptotyphlopids, which are thought to use existing burrows rather than digging their own [@herrelBurrowing2021]. In this regard, the Australian typhlopid genus *Anilios* is quite exceptional because there is great variation in body shape [@tiatragulMorphological2024]. Species that evolved more narrow-and-long bodies may be converging on the behaviour with other scolecophidians that have similar phenotypes (i.e., leptotyphlopids). However, more detailed species-level microhabitat observations are required to determine if these species are actually taking advantage of these environments in the ways we speculate.

>>>>

Evolution of the axial skeleton is fundamental to vertebrate body plan evolution, facilitating body form changes and influencing modes of locomotion in mammals [e.g., @taewcharoenSmall2024;@galisFast2014], reptiles [e.g., @vandammeSpeed2002;@molnarMorphological2023], amphibians [e.g., @bonettEvidence2017;@adlerEvolutionary2022] and fish [e.g., @yamadaAdaptive2009;@forkerBendy2021]. Axial skeleton evolution is often studied by counting the number of vertebrae because the developmental mechanisms of segmentation are well characterised. Essentially vertebrae begin as somites, epithelial blocks of mesoderm, which form rhythmically during embryogenesis where the speed of this process relates to the size and number of somites, to produce what will become the axial skeleton [@gomezControl2008;@dequeantSegmental2008;@gomezDevelopmental2009]. Elongate animals are expected to show a positive correlation between maximum adult body length and number of vertebrae, known as pleomerism [@lindseyPleomerism1975]. While pleomerism was considered the norm [@lindseyPleomerism1975;@wakeComparative1966;@pollyTesting2001], more recent studies have identified exceptions to the pleomerism rule in moray eels [@reeceEvolutionary2013], large species of snakes [@headDissociation2007;@lindellEvolution1994], and salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].

The lack of pleomerism in some species suggests it is possible to disassociate segment production with body growth, and thus reveals the developmental processes underlying evolutionary patterns. Another clue to disassociation in reptiles can be found in the relationship between the number of ventral scales and precloacal vertebrae – a correlation that is 1:1 in many snakes [@alexanderPattern1966] and so ventral scales have often been used as a proxy for vertebrae number [@lindellEvolution1994;@shineVertebral2000]. However, this 1:1 correlation is absent in file snakes, *Achrochodus*, and some blindsnakes, Typhlopidae, [@alexanderPattern1966;@headDissociation2007], as well as some sea snakes of the genus *Hydrophis* [@vorisDermal1975], suggesting these taxa do not conform to pleomerism.
Snakes are therefore a good model to study axial evolution because they have great diversity in vertebrae number and a wide range of adult body sizes and shapes but without a limb skeleton to confound signals of locomotor adaptation. Recent studies have indicated the number of vertebrae varies predictably with ecological diversification in elapid snakes [@sherrattVertebral2019;@hamptonInterspecific2019;@sherrattMacroevolution2022] and colubrid snakes [@nash-hahnModification2024]. 
The number of vertebrae has also been shown to influence snake locomotion, whereby the number of vertebrae negatively correlates with speed travelled in garter snakes [@kelleyEffects1997]. Despite these recent advances, there have been no study that have quantified the number of vertebrae across all the species in a diverse assemblage of fossorial snakes to test for the variation and potential ecological drivers.

<!-- [Specific background]-->
Scolecophidians, commonly known in English as "blindsnakes" and "threadsnakes", are characterised by a reduced visual system and highly specialised phenotypes adapted for a strictly fossorial existence [@mirallesMolecular2018;@strongDeconstructing2021;@gowerEyetranscriptome2021]. Scolecophidians comprise five paraphyletic families, including the "core scolecophidians" — a monophyletic clade consisting of Typhlopidae (425 spp), Leptotyphlopidae (143 spp), Gerrhopilidae (29 spp), and Xenotyphlopidae (1 spp) — and the paraphyletic Anomalepididae (20 spp), which is sister to all other "typical snakes" (Alethinophidia) [@mirallesMolecular2018; @burbrinkInterrogating2020; @uetzQuarter2021]. Although all blindsnakes are fossorial, body size and shape vary significantly among species [@tiatragulMorphological2024; @shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993]. Leptotyphlopids are generally considered more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@hedgesLower2008;@webbDietary2001;@webbPreysize1993;@iturriagaNew2021;@tiatragulMorphological2024]. These differences in body size and proportions determine how much push force the snakes can generate when burrowing, which led to the idea that leptotyphlopids generally use existing burrows while typhlopids create their own burrow [@herrelBurrowing2021].

@@

We collected body size measurements and vertebral counts from museum specimens examined. Body measurements including snout-to-vent length (SVL), tail lengths (TL) and midbody diameter (MBD) were obtained from @tiatragulMorphological2024. SVL and TL were measured to the nearest 1 mm on a ruler. MBD was measured using a digital vernier calliper (ABS Coolant Proof Caliper, Mitutoyo, Japan) to the nearest 0.1 mm. Total body length (TBL) is the sum of SVL and TL. To quantify body shape for each specimen, we calculated the aspect ratio — also referred to as elongation ratio [@wardConvergent2004] — which is the quotient of TBL and MBD.

@@

We used the Fiji distribution of ImageJ2 v.1.54j [@schindelinFiji2012;@ruedenImageJ22017] to digitally visualise X-ray images and counted the number of vertebrae using the Cell Counter module. 
<!-- Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023].  -->
For each specimen we calculated vertebrae ratio by dividing the total number of vertebrae by TBL.

@@

All statistical analyses were performed in R Statistical Environment v 4.4.0 [@rcoreteamLanguage2024]. To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Brownian Motion implemented in the R package *`geomorph`* v.4.0.6 [@adamsGeomorph2023]. First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes. We did this by regressing the species maximum number of total vertebrae against maximum TBL. Next, we tested whether the species mean aspect ratio is correlated with mean vertebrae ratio (mean vertebrae/mean TBL) and/or mean vertebrae number. Finally, we tested if variation in vertebrae ratio and vertebrae number is correlated with maximum soil bulk density and mean annual temperature. In all PGLS analyses, statistical significance was assessed by permutation, using 1000 iterations. To visualise the relationship between vertebrae number and intracolumnar vertebral length, for each of the representative species, we plotted the mean vertebrae size (from three repeated measurements) against the vertebra number. Each vertebrae profile was overlaid with a fitted 4° polynomial curve [e.g., @sherrattMacroevolution2022].

@@

In a related study, @tiatragulMorphological2024 found that Australian blindsnake species with smaller absolute body size generally occupy areas that are warmer regions of Australia than while larger species tend to occupy cooler regions. Here, we observe that vertebrae ratio (raw vertebrae number/SVL) is positively correlated with temperature, indicating species that have more vertebrae than expected from their body size are found in warmer places. In his work on rattlesnakes, @klauberRattlesnakes1982  noted that smaller-sized subspecies often retain the same number of vertebrae as their larger closest relative, which result in the smaller subspecies having relatively greater vertebrae ratio. An explanation for this pattern is that size is often under greater selection than vertebral number, and there is negligible disadvantage for a snake in having too many vertebrae for their body size [@lindellEvolution1994]. This makes sense when we also consider the negative relationship between vertebrae ratio and maximum soil bulk density. A greater number of (shorter) vertebrae for body size may afford greater flexibility [@jayneKinematics1986;@bergmannVertebral2012;@johnsonAdaptive1955] that enables these species to navigate in pre-existing burrows or termite galleries, typical across northern Australia, without exerting force to dig their own burrow. @tiatragulMorphological2024 also found species that occupy areas with more difficult-to-penetrate soil are usually narrow-and-long rather than robust-and-wide species that occupy areas with easier-to-penetrate soil. Narrow-and-wide phenotypes are common among leptotyphlopids, which are thought to use existing burrows rather than digging their own [@herrelBurrowing2021]. In this regard, the Australian typhlopid genus *Anilios* is quite exceptional because there is great variation in body shape [@tiatragulMorphological2024]. Species that evolved more narrow-and-long bodies may be converging on the behaviour with other scolecophidians that have similar phenotypes (i.e., leptotyphlopids). However, more detailed species-level microhabitat observations are required to determine if these species are actually taking advantage of these environments in the ways we speculate.
